---
title: "R concepts, MiCM Fall 2022 Workshop"
author: "Alex Diaz-Papkovich"
date: "04/11/2022"
output: html_document
---

If you ever want to get rid of a variable or object, use the `rm` command to delete it. To remove everything, use `rm(list=ls())`. Your code will be safe as long as you've saved it, but any variables, data frames, plots, environment objects, etc, will be removed. Libraries/packages that you've loaded will be unloaded --- they will still be installed, but they will not be active and you will have to use `library([package name])` to reload it.

Small tricks:
* Wrap a command in `()` to output its result in the console, e.g. `(x <- c(1,2,3))` will assign values to `x` and also print it.

We will begin by assigning values to a vector.

```{r Basic variable assignments}
# The "<-" and "=" operators are equivalent
# This code creates a vector of numeric variables and prints it
x <- c(1,2,3)
print("This is my variable:")
print(x)

x = c(1,2,3)
print("This is my variable:")
print(x)
```

We will explore the main types of variables:

```{r The multiple types of variables}
# Numeric
numeric_vector <- c(1,2,3,4,5)
print(numeric_vector)

# Character
character_vector <- c("one","one","three","four","five")
print(character_vector)

# Logical
logical_vector <- c(T,F,T,T,F)
print(logical_vector)

# Factor
# Note the "levels" when you print them
factor_vector <- as.factor(c("Toronto","Toronto","Montreal","Montreal","Ottawa","Montreal","Toronto","Montreal"))
print(factor_vector)

# Examine the levels of a factor
levels(factor_vector)

as.numeric(factor_vector) # Factors have an underlying number related to their label
as.numeric(character_vector) # Characters do not
```

We can also convert one type of variable to another, if it is appropriate

```{r Convert variable types}
character_of_numbers <- c("1","2","3","4")
converted_characters <- as.numeric(character_of_numbers)
print(character_of_numbers)
print(converted_characters)
```

Lists can contain multiple variable types. Sometimes, you will want to add a row to a data frame. One way is to append a list to it.

```{r Example of a list}
# This list contains several types of variables
my_list <- list(1,"2",as.factor("ten"),FALSE)
# We can look at the structure of the list
str(my_list)

# Vectors do not support this function
my_vector <- c(1,"2",as.factor("ten"),FALSE)
str(my_vector)
```

Finally, we look at a data frame.
```{r Example of a data frame}
# Create vectors that will contain our data
# We will create the columns
FirstCol <- c("Toronto","Toronto","Montreal","Halifax","Gatineau","Gatineau","Halifax","Toronto","Vancouver")
SecondCol <- c(21,24,25,33,20,18,19,20,20)
ThirdCol <- c("M","F","M","M","F","F","X","F","M")

# Create the data frame
my_data <- data.frame(FirstCol,SecondCol,ThirdCol)

# These column names are not very informative, so we should change them
colnames(my_data) # Look at the column names

# Re-assign the column names
colnames(my_data) <- c("City","Age","Sex")

# Let us select a specific column, in this case Age
age_vector <- my_data$Age
print(age_vector)

# Let's add a row of data
new_row <- list(as.factor("Toronto"),20,"F")
data_with_new_row <- rbind.data.frame(my_data, new_row)
```

### Exercises

Exercise 1: Create a data frame from the vectors provided. Name the columns "age", "sex", "location"

```{r Exercise 1}

v1 <- c("20","27","26","24","27","18","22","23","24","29")
v2 <- c("M","M","F","F","M","M","F","F","M","F")
v3 <- c("Ottawa","Ottawa","Montreal","Vancouver","Toronto","Halifax","Toronto","Calgary","Montreal","Toronto")

```

Exercise 2: Convert the columns for age, sex, and location to `numeric`,`factor`, and `factor`, respectively.

```{r Exercise 2}
```

Exercise 3: Half of the individuals have been assigned a treatment, and another half a placebo. Add a column of `logical` values called "treatment" using `cbind` or `cbind.data.frame`.

```{r Exercise 3}
v4 <- c(rep(TRUE,5), rep(FALSE,5))
```

Exercise 4: Add an observation to your data. (Optional)

```{r Optional exercise 4}
```